---
title: "Rgronomist_nlme"
author: "Rafael Martinez-Feria"
date: "11/10/2017"
output: html_document
---

## Fitting non-linear mixed effects models using `nlme` package. 

I will be following the approach described by [Archontoulis and Miguez (2015)](https://dl.sciencesocieties.org/publications/aj/articles/107/2/786)

```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(nlme) # for curve 
library(ggplot2) # for nice plotting 
```

We will be working with the Oats data set included in the nlme package. 

```{r}
oats <- as.data.frame(nlme::Oats)
```

From the `nlme` documentation:

> These data have been introduced by Yates (1935) as an example of a split-plot design. The treatment structure used in the experiment was a 3 x 4 full factorial, with three varieties of oats and four concentrations of nitrogen. The experimental units were arranged into six blocks, each with three whole-plots subdivided into four subplots. The varieties of oats were assigned randomly to the whole-plots and the concentrations of nitrogen to the subplots. All four concentrations of nitrogen were used on each whole-plot.

```{r}
# I don't like the units so I convert to kg/ha
oats$nitro <- round(oats$nitro*100*1.12,0)
oats$yield <- oats$yield*32*1.12
```

We need to save as `groupedData()` to help us during the fitting process

```{r}
oats <- groupedData(yield ~ nitro | Block,
                    data = oats, 
                    units = list(x = "N rate (kg/ha)",
                                 y = "Yield (kg/ha)"))

plot(oats) # vizualize data 
```

## A word about mixed-effect models

How is a mixed model different than a linear fixed model?

```{r}
# Fit simple linear model without random effects
fit.lm <- lm(yield~Variety*nitro, data = oats)
anova(fit.lm)
```

```{r}
summary(fit.lm)
```

```{r}
coefficients(fit.lm)
```

In a mixed model we use the grouping structure of the data (here being the split-plot) to assing a random effect (here being the plot).

```{r}
# Fit a linear mixed-effects model with subplot as a random effect
fit.lme <- lme(yield~Variety*nitro, random = ~1|Block/Variety,# why not random = ~1|Block/Variety/nitro?
               data = oats)  
anova(fit.lme)
```

```{r}
summary(fit.lme)
```


```{r}
coefficients(fit.lme)
```

```{r}
expand.grid(Block = unique(oats$Block),
            Variety = unique(oats$Variety),
            nitro = 1:70) -> newOats

newOats$predict <- predict(fit.lme, newdata = newOats)
newOats <- merge(newOats, data.frame(oats),all = T)

ggplot(newOats,
       aes(y= predict, x = nitro, colour = Variety)) +
  geom_line() + 
  geom_point(aes(y=yield)) + 
  facet_wrap(~Block, ncol = 3) + 
  theme(legend.position = "top")
```
```{r}
ggplot(newOats,
       aes(y= predict, x = nitro, colour = Block)) +
  geom_line() + 
  geom_point(aes(y=yield)) + 
  facet_grid(.~Variety)
```

## Why a non-linear model? 

Many biological processes are not linear. For intance, can we expect for yield
to increase linearly with N fertilizer indefinitively?

**Non-linear hypothesis:** There is a level where N that no longer increases yield.

Here we use a linear plateau model: 
```{r}
linearPlateau <- function(x,a,b,c) a + b * (x - c) * (x <= c)
```

The nlsList function uses the group data estrcuture to fit a curve for each group (i.e. block)

```{r}
fit.nls <- nlsList(yield ~ linearPlateau(nitro,a,b,c),
                   start = list(a = 30, b = 50, c = 60),
                   data = oats)

summary(fit.nls)
```
```{r}
plot(fit.nls)
```

Now we define our random effects on the model parameters. Here nlme knows the data structure we passed as groupedData() so it knows the random effect is the block. 


```{r}
fit.nlme <- nlme(fit.nls, random = pdDiag(a + b + c ~ 1))
plot(ranef(fit.nlme))

```

```{r}
random.effects(fit.nlme) # look at the scale for b and c
```

```{r}
anova(fit.nlme)
```

```{r}
coef(fit.nlme)
```

```{r}
(fe <- fixed.effects(fit.nlme))
```

Now we can ask: Is there an effect of variety?
```{r}
update(fit.nlme, 
       fixed = list(a + b + c ~ Variety),
       start = c(fe[1],5,5, #Here is where you'll spend most of the time
                 fe[2],0,0,
                 fe[3],0,0)) -> fit.nlme2

anova(fit.nlme2)

```

```{r}
coef(fit.nlme2)
```

```{r}
plot(augPred(fit.nlme2,level = c(0,1)))
```

# Final plot 
```{r}
expand.grid(Block = unique(oats$Block),
            Variety = unique(oats$Variety),
            nitro = 1:70) -> newOats

newOats$predict <- predict(fit.nlme, newdata = newOats)
newOats <- merge(newOats, data.frame(oats),all = T)

ggplot(newOats,
       aes(y= predict, x = nitro, colour = Variety)) +
  geom_line() + 
  geom_point(aes(y=yield)) + 
  facet_wrap(~Block,ncol = 3)

```


```{r}
ggplot(newOats,
       aes(y= predict, x = nitro, colour = Block)) +
  geom_line() + 
  geom_point(aes(y=yield)) + 
  facet_grid(.~Variety)
```